Explainable AI with Probabilistic Graphical Models

### 2019-11-13 literature-review/explainability.tex
Thesis complete

\section{Explainability} \label{sec:explainability}
\citet{Doran2018}, based on a frequency analysis of explanation terms within documents from relevant research communities, highlight how different circles have different approaches to the concept of explainability and how, even within the same group, terms are used interchangeably.
In particular, they note the overloading of the notion of \enquote{explaination} with that of \enquote{interpretability}; a concept that is often defined within the xAI community as necessary for, but distinct from, explainability.
The use of \enquote{interpretable}, as signifying the property belonging to a system whose inner workings are accessible, can be found, for example, in the recent paper by \citet{gilpin2018explaining}.
In other recent works the two terms are conflated, for example by \citet{mittelstadt2019explaining}, \citet{guidotti2018survey} and in the influential work by \citet{doshi2017towards}.
This seems to prove the point made by \citet{Lipton2016} in the widely-cited paper \enquote{The Mythos of Model Interpretability}, that \enquote{the task of interpretation appears underspecified. Papers provide diverse and sometimes non-overlapping motivations for interpretability, and offer myriad notions of what attributes render models interpretable}.

Most works, even those that blend the notions of interpretability and explainability, seem to agree on the end-goal that implementing such a concept should have; that is, to \enquote{summarize the reasons for neural network behavior, gain the trust of users, or produce insights about the causes of their decisions} \citep{gilpin2018explaining} by being able to \enquote{explain or to present in understandable terms to a human} \citep{doshi2017towards}.
Where the consensus diverges, is in defining what constitutes an explanation and in fixing the desiderata that it may have.
\citet{mittelstadt2019explaining} identify within the literature, two broad classes of interpretability/explainability: \textit{transparency} (also called \textit{ante-hoc} explainabilility) and \textit{post-hoc} interpretability.
The former type deals with the internal workings of a system while the latter applies to its external behaviour.
\citet{Lipton2016} identifies three explanations that can make a model transparent: a mechanistic understanding of the workings of the system in its entirety, of the individual components or of the algorithm.
A system may be made post-hoc interpretable by way of, among others, natural language explanations, visualisations or interactive interfaces.
These methods often do not precisely clarify the exact working of a model, but \enquote{they may nonetheless confer useful information for practitioners and end-users of machine learning}.
\citet{Biran2017} notes how the transparent or \textit{white-box} paradigm was sufficient for classic rule-based models but - with the advent of contemporary machine learning models - is no longer useful.
They argue that it is nowadays unreasonable to expect that any domain expert be able to understand a prediction if they are not also a machine learning specialist.
To address this issue, they propose a Natural Language Generation system; that is, a post-hoc explanation in the categorisation by \citet{mittelstadt2019explaining}.

A widely-recognised feeling, tightly connected with the already recognised lack of shared working definitions, seems to be that researches of Explainable AI are ignoring the enormous corpus of existing models in philosophy, psychology, cognitive and social sciences and human-computer interaction.
This feeling of disconnect is echoed by \citet{gilpin2018explaining} who points out how philosophical texts have long debated what constitutes an explanation, by \citet{mittelstadt2019explaining} who explicitly says how \enquote{many different people, be they lawyers, regulators, machine learning specialists, philosophers, or futurologists, are all prepared to agree on the importance of Explainable AI [...] very few stop to check what they are agreeing to, and to find out what Explainable AI means to other people involved in the discussion}.
The fact that Explainable AI researchers seem to be intent on \enquote{reinventing the wheel} is stated most strongly by \citet{miller2018explanation}, whose paper \enquote{Explanation in Artificial Intelligence: Insights from the Social Sciences} is based on the premiss that \enquote{most of the research and practice in this area seems to use the researchers' intuitions of what constitutes a `good' explanation} and argues for the adoption of the existing research in the social sciences.
The author's views are well summarised by the position xAI is set to occupy in Figure \ref{fig:xai-position}.
The feeling is that Explainable AI researchers are terming their methods \enquote{explanation} based on purely personal views and are thus building explanations that only work for themselves; in other words, \enquote{the inmates are running the asylum} \citep{Miller2017}.
The following quote by \citet{guidotti2018survey} seem to perfectly sum up the state of the research in the field: 
\begin{quotation}
	It is evident that the research activity in this field completely ignored the importance of studying a general and standard formalism for defining an explanation, identifying which are the properties that an explanation should guarantee, e.g., soundness, completeness, compactness and comprehensibility. Concerning this last property, there is no work that seriously addresses the problem of quantifying the grade of comprehensibility of an explanation for humans, although it is of fundamental importance.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.5\textwidth]{literature-review/images/xai-position}}
\caption{Venn diagram showing where the field of xAI should \textit{ideally} be positioned \citep{miller2018explanation}}
\label{fig:xai-position}
\end{figure}

To remedy to this state of affairs, there have been a number of works, such as that by \citet{doshi2017towards}, that attempted to define what an explanation means and to reach consensus on it.
The most compelling attempt is in a paper by \citet{Doran2018} (already mentioned in Section \ref{sec:intro-context}) where the authors try to synthesise the current state of affairs into a taxonomy of models:
\begin{itemize}
	\item \textit{Opaque systems}: these are systems that offer no insight into their internal workings that transform the inputs symbols, usually real-world data, into some output, for example labels that classify the data or predictions of unseen cases. 
		All closed-source algorithms fall under this definition.
	\item \textit{Interpretable systems}: this is the vastest category, as the characteristic of these systems is \textit{transparency} i.e., their inner workings are accessible but the onus of comprehensibility falls completely onto the user.  
		The classical example is that of Neural Networks where the mapping from inputs to outputs is inspectable by the user who can, theoretically and depending on her skill, interpret them.
		In the case of Neural Networks this may be a daunting task due to their distributed and non-symbolic nature; in other classes of Machine Learning models, for example Bayesian Networks, the task may be easier as discussed in Section \ref{sec:explainability-in-bayesian-networks}.
	\item \textit{Comprehensible systems}: systems in this category emit additional symbols together with their outputs with the explicit intent of giving the user the means to interpret and understand the automated decisions.
		The additional symbols may be visualisations, natural-language text or any other means of demystifying the output.  
		These extra symbols would need to be graded based on the user's expertise, as comprehension is a property that involves both man and machine.
		An example of such a system would be an image classifier that together with its output also highlighted the parts of the image it used to make its decision.
	\item \textit{Explainable systems}: the highest level in the taxonomy includes those models that emit an explicit \textit{explanation} i.e., a human-understandable line of reasoning.
\end{itemize}
It is recognised that comprensibility depends not only on the system's characteristics, but also on the user's ability and knowledge, thus implicitly accepting the view that xAI should learn from the social sciences.
\textit{Comprensibility} and \textit{interpretability} are considered separate concepts as comprehension requires transparency but interpretation does not, given that the user may reason only over the emitted extra symbols.
This notion of comprensibility is expanded into that of \textit{real} explainability, that is based on a notion of \enquote{ability to formulate, for the user, a line of reasoning that explains the decision making process of a model using human-understandable features of the input data}.

\section{Importance of Explainability} \label{sec:importance-of-explainability}
As noted by \citet{edwards2018enslaving}, \enquote{businesses and governments are increasingly deploying machine learning (ML) systems to make and support decisions that have a crucial impact on everyday life} so, as \citet{gilpin2018explaining} say, \enquote{it becomes necessary for these mechanisms to explain themselves}.
This feeling of urgency and purpose is echoed throughout the reviewed literature; it seems, that even if researchers and the field as a whole cannot agree on a definition of explainability (as discussed in Section \ref{sec:explainability}), there is a keen awareness on the need for models to be explainable.
This intense urge to define explainability and, at the same time, try and create models exhibiting this property, may be counterproductive as the field risks fragmenting into a series of diverging strands, as noted by \citet{abdul2018trends} and visualised in Figure \ref{fig:xai-citation-network}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.8\textwidth]{literature-review/images/xai-citation-network}}
\caption{Citation network that is emblematic in showing the breadth of research strands in the field of xAI \citep{abdul2018trends}}
\label{fig:xai-citation-network}
\end{figure}

There are a myriad of reasons that are brought forth as a justification for the development of Explainable AI, and we will review these in the following paragraphs, but it seems timely to start with one in particular: the need introduced by the European Union's broad General Data Protection Regulation (GDPR).
The GDPR was approved by the European Parliament in 2016 and came into effect in 2018.
More than one author cites an urgency to conform to this regulation (for example \citet{doshi2017towards}, \citet{gilpin2018explaining}), most likely referring to Article 22 of the regulation that, supposedly, mandates for a \enquote{right to an explanation} of algorithms.
While algorithmic explainability is undoubtedly a commendable end-goal, it may be the case that this particular reason to strive for it be a false one.
\citet{edwards2018enslaving} posit that Article 22 of the GDPR actually does not contain the publicised right to an explanation but is \enquote{merely a right to stop processing unless a human is introduced to review the decision on challenge} and as the authors point out, there are, nowadays at least, very few systems without a human in the loop.
Secondly, there is no mandate for the \enquote{explanation} to be human-understandable, so the obtained result may actually be no explanation at all.
If the analysis of \citet{edwards2018enslaving} were correct, then the urgency advocated by many researchers on the grounds of conforming to the GDPR would turn out, in reality, to be based on no valid reason.

A second motive brought forth for the necessity of explainability, for example by \citet{gilpin2018explaining} and \citet{abdul2018trends}, is that comprehensible models are much more likely, or even necessary, to engender users' trust.
While this may very probably be the case, no motives are given for why this should be the purpose of Explainable AI and not just a desirable by-product of obtaining explainability. 
A reasoning for why trust may stem from explainability can be found in \citep{Kyrimi2016} where it is claimed that \enquote{the lack of trust may be due to the difficulty of understanding how a prediction is inferred from the given data. As Aristotle wrote `we do not have knowledge of a thing until we have grasped its why, that is to say, its explanation'. Hence, explaining a model's reasoning - its inference - could increase trustworthiness.}

Other authors, for example, \citet{doshi2017towards} and \citet{guidotti2018survey}, frame the issue as one of moral necessity.
One need not look far to find examples of ML models displaying covert bias or making decisions we would regard as unethical; a more in-depth investigation would reveal that this was the case even before the popularisation of \enquote{black box} models as are Deep Neural Networks.
\citet{guidotti2018survey} give a reasonably comprehensive list of classic cases that show the risks inherent in not having comprehensible AI.
The oldest of these dates back to the 1970s and 1980s and tells of a system used to screen job applicants that, even though programmed to ignore people's ethnicity, was still seen to discriminate against minorities.
In the same vein but much more recent, the American Military discovered that their computer vision system, that was developed to differentiate between enemy and friendly tanks automatically, had poor accuracy because it had learned to use the background information of the test set photos instead of the pixels representing the actual tanks.
Both these cases exemplify how an algorithm may make \enquote{wrong} inferences based on spurious or latent information that was already in the data set but that no human could have imagined being relevant.
Other failures epitomise how a model may learn our own social prejudices; for example, a recent Princeton study \citep{caliskan2017semantics} proved how models trained on web text corpora showed marked biases (towards race, gender ...) that reflected the ones present in our own society.
Based of their findings, the authors went as far as to suggest that transparency would not be enough to uproot biases, as the very \textit{semantics} of language reflects prejudice latent in our culture.

The driving motivation and sense of urgency present in all of the reviewed literature is quite certainly tied to the renewed interest and applicability of AI.
\citet{Preece2018} claim that the interest for explainability is naturally linked to that in AI; if this were true, then it would confirm that a need for transparency is implicit in the field itself.
This would validate the assertion made by \citet{doshi2017towards}, that the need for an explanation stems from an \textit{incompleteness in the formalisation}.
What is meant by this is that optimising for certain objectives may introduce an \textit{unquantifiable bias} into the system, that is very different than mere \textit{uncertainty}.
Mere uncertainty can be rigorously quantified, formalised and reasoned upon by using Probability Theory; unquantifiable bias is the result of an \textit{incompleteness in formalisation} of the problem that the ML system is tasked with.
This is the case when a system is coded, for example, to pursue soft-objectives such as \textit{ethics}; such an end-goal may be too abstract and nebulous to be completely formalised.
It might also be the case that a particular objective is far too complex for all its possible outcomes to be exhaustively enumerated.
A good characterisation of objectives leading to incompleteness could be given by applying the concept of \enquote{wicked problem} introduced by \citet{Rittel1973} when analysing the issues arising in social planning.
The first defining characteristic of a wicked problem is that the mere definition of the problem at hand is the wicked problem itself, since even its description is dependent on one's idea for solving it; there is no definite locus that one can point to as the source of the problem. 
Other defining features are the absence of stopping and objective evaluation criteria and the fact that each solution to a wicked problem is essentially \enquote{one-shot} and unique.  
Even the set of possible solutions and causes are not predetermined and non-stationary. 
The authors noticed how these problems presented a series of defining characteristics, that they were able to generalise into the notion of wicked problem.
A classic example are the issues that arise when trying to solve a social planning problem: various parties will have competing objectives and different ideas to obtain them, so no unique best solution is possible; \enquote{at best, they are only re-solved} \citep{Rittel1973}. 
These situations generally tend to arise when dealing with human values, as there is always a degree of ethical relativism that makes it difficult to develop and evaluate solutions.  
\citet{Lipton2016} is also aware of this and states that \enquote{the demand for interpretability arises when there is a mismatch between the formal objectives of supervised learning and the real world costs in a deployment setting}.
In the presence of such unquantifiable uncertainty, that is unavoidable in many of the important applications of AI, requiring the resulting model to be open to inspection would make the \enquote{gaps in problem formalization visible to us} and thus enable us to apply our best human judgement to evaluate them and their consequences \citep{doshi2017towards}.
No wicked problem has a solution that is either true or false, but only good or bad; necessarily, the only judge for this can be a human.

\section{Evaluation of Explainability} \label{sec:evaluation-of-explainability}
As concluded in Section \ref{sec:importance-of-explainability}, the fact that ML models operate on incomplete assumptions makes it a necessity to have some form of evaluation of their performance.
As \citet{Lipton2016} states, \enquote{it turns out that many situations arise when our real world objectives are difficult to encode as simple real-valued functions} and this could lead to evident difficulties in optimising with respect to soft, but of paramount importance, concepts such as ethics and legality.
Being able to evaluate an automated explanation lets us \enquote{serve those objectives that we deem necessary but struggle to model formally}.

Unfortunately, the finding outlined in Section \ref{sec:explainability}, namely there being no consensus on the definition of explainability also necessarily entails that there is no agreed-upon methodology to evaluate such a property.
\citet{doshi2017towards} note as much when they comment \enquote{unfortunately, there is little consensus on what interpretability in machine learning is and how to evaluate it for benchmarking}.
This makes perfect sense because trying to evaluate something without first having defined it, is worse than trying to hit a moving target.
Once again, the feeling among many authors is that \enquote{inmates are running the asylum}.

Some authors have tried to put some order in the barrage of proposed methods; \citet{doshi2017towards} provide one of the most compelling attempts.
These authors set out to outline a taxonomy, having noted a \enquote{lack of rigour} and how current interpretability approaches usually fall into two categories: interpretability in the context of an application and interpretability via a quantifiable proxy.
 The former approach assumes that \enquote{if the system is useful in either a practical application or a simplified version of it, then it must be somehow interpretable}; the latter sees researchers claim that a model class is interpretable and then present algorithms to optimise within that class.
 In their words, both classes rely on a notion of \enquote{you'll know it when you see it}.
 The taxonomy the authors propose is laid out in Figure \ref{fig:xai-taxonomy} and borrows from methods already standard in human-computer interaction and visualisation; the guiding ideal is that \enquote{evaluation of applied work should demonstrate success in the application} and thus the best kind of evaluation is the one that involves humans the most:
 \begin{itemize}
  \item \textit{Functionally-grounded Evaluations}: at the lowest level of their taxonomy are those methods that require no human-in-the-loop and evaluate the explanation quality of a system by using some proxy measure; the advantage is the low cost, but the tradeoff is a lack of specificity.
 A proxy measure that has already been human-validated, for example a decision tree, a set of rules or a linear model \citep{guidotti2018survey} as regards its explainability, may be an appropriate measure for more exotic systems.
 \item \textit{Human-grounded Evaluation}: the second level of evaluation involves humans, albeit not domain expert ones, carrying out simplified versions of the target application; this kind of setup enables the testing of more general notions of explainability.
 \item \textit{Application-grounded evaluation}: this is considered the gold standard evaluation; the authors claim that there is no better way to evaluate explanability that having a domain expert test it in the context of a real task.
In their words, \enquote{the best way to show that the model works is to evaluate it with respect to the task}: \enquote{for example, a visualization for correcting segmentations from microscopy data would be evaluated via user studies on segmentation on the target image task; a homework-hint system is evaluated on whether the student achieves better post-test performance.  Specifically, we evaluate the quality of an explanation in the context of its end-task, such as whether it results in better identification of errors, new facts, or less discrimination}.
\end{itemize}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{literature-review/images/xai-taxonomy}}
\caption{Taxonomy of methods for the evaluation of explanations \citep{doshi2017towards}}
\label{fig:xai-taxonomy}
\end{figure}

\citet{guidotti2018survey} outline a taxonomy developed along a different axis; specifically they identify: 
\begin{itemize}
  \item \textit{methods to explain black box models};
  \item \textit{methods to explain black box outcomes};
  \item \textit{methods to inspect black boxes and methods to design transparent boxes}.
\end{itemize}
An essential factor that the authors identify is that \textit{evaluation is a graded notion}; as also noted by \citet{gilpin2018explaining}: different users, with different expertise and background, may rate the quality of the same explanations very differently.
Another point they make - that is quite novel - is that time should also be part of an evaluation; depending on the available time a user may prefer a more straightfoward or more elaborate explanation.
They end by noting that very few works take user expertise and the time taken to understand the proposed explanation into account.

>>>>

\section{Explainability} \label{sec:explainability}
\citet{Doran2018}, carried out a frequency analysis of explanation terms within documents from relevant research communities.
They highlight how different circles have different approaches to the concept of explainability and how, even within the same group, terms are used interchangeably.
In particular, they note the overloading of the notion of \textit{explanation} with that of \textit{interpretability}; a concept that is often defined within the xAI community as necessary for, but distinct from, explainability.
The use of \textit{interpretable}, as signifying the property belonging to a system whose inner workings are accessible, can be found, for example, in the recent paper \citep{gilpin2018explaining}.
In other recent works the two terms are conflated, for example in \citep{mittelstadt2019explaining}, \citep{guidotti2018survey} and in the influential work by \citet{doshi2017towards}.
This seems to prove the point made by \citet{Lipton2016} in the widely-cited paper \enquote{The Mythos of Model Interpretability}, that \enquote{the task of interpretation appears underspecified. Papers provide diverse and sometimes non-overlapping motivations for interpretability and offer myriad notions of what attributes render models interpretable}.

Most works, even those that blend the notions of interpretability and explainability, seem to agree on the end-goal that implementing such a concept should have; that is, to \enquote{summarize the reasons for neural network behavior, gain the trust of users, or produce insights about the causes of their decisions} \citep{gilpin2018explaining} by being able to \enquote{explain or to present in understandable terms to a human} \citep{doshi2017towards}.
Where the consensus diverges, is in defining what constitutes an explanation and in fixing the desiderata that it may have.
\citet{mittelstadt2019explaining} identify two broad classes of interpretability/explainability within the literature: \textit{transparency} (also called \textit{ante-hoc} explainability) and \textit{post-hoc} interpretability.
The former type deals with the internal workings of a system while the latter applies to its external behaviour.
\citet{Lipton2016} identifies three explanations that can make a model transparent: a mechanistic understanding of the workings of the system in its entirety, of the individual components or of the algorithm.
A system may be made post-hoc interpretable by way of natural language explanations, visualisations or interactive interfaces, among others.
These methods often do not precisely clarify the exact working of a model, but \enquote{they may nonetheless confer useful information for practitioners and end-users of machine learning}.
\citet{Biran2017} note how the transparent or \textit{white-box} paradigm was sufficient for classic rule-based models but - with the advent of contemporary machine learning models - is no longer useful.
They argue that it is nowadays unreasonable to expect that any domain expert be able to understand a prediction if they are not also a machine learning specialist.
To address this issue, they propose a \textit{natural language generation system}; that is, a \textit{post-hoc} explanation in the categorisation by \citet{mittelstadt2019explaining}.

A widely-recognised feeling, closely connected with the already identified lack of shared working definitions, seems to be that researchers of explainable AI are ignoring the enormous corpus of existing work in the fields of philosophy, psychology, cognitive and social sciences and human-computer interaction.
This feeling of disconnect is echoed by \citet{gilpin2018explaining} who point out how philosophical texts have long debated what constitutes an explanation and by \citet{mittelstadt2019explaining} who explicitly say how \enquote{many different people, be they lawyers, regulators, machine learning specialists, philosophers, or futurologists, are all prepared to agree on the importance of explainable AI [...] very few stop to check what they are agreeing to, and to find out what explainable AI means to other people involved in the discussion}.
The fact that explainable AI researchers seem to be intent on \enquote{reinventing the wheel} is stated most strongly by \citet{miller2018explanation}, whose paper \enquote{Explanation in Artificial Intelligence: Insights from the Social Sciences} is based on the premiss that \enquote{most of the research and practice in this area seems to use the researchers' intuitions of what constitutes a `good' explanation} and argues for the adoption of the existing research in the social sciences.
The author's views are well summarised by the position xAI is set to occupy in Figure \ref{fig:xai-position}.
The feeling is that explainable AI researchers are calling their methods an \textit{explanation} based on purely personal views and are thus building explanations that only work for themselves; in other words, \enquote{the inmates are running the asylum} \citep{Miller2017}.
The following quote from \citet{guidotti2018survey} seem to perfectly sum up the state of the research in the field: 
\begin{quotation}
	It is evident that the research activity in this field completely ignored the importance of studying a general and standard formalism for defining an explanation, identifying which are the properties that an explanation should guarantee, e.g., soundness, completeness, compactness and comprehensibility. Concerning this last property, there is no work that seriously addresses the problem of quantifying the grade of comprehensibility of an explanation for humans, although it is of fundamental importance.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.5\textwidth]{literature-review/images/xai-position}}
\caption{Venn diagram showing where the field of xAI should \textit{ideally} be positioned \citep{miller2018explanation}.}
\label{fig:xai-position}
\end{figure}

To remedy this state of affairs, there have been a number of works, such as \citep{doshi2017towards}, which have attempted to define what an explanation means and to reach a consensus on it.
The most compelling attempt is in a paper by \citet{Doran2018} (already mentioned in Section \ref{sec:intro-context}) where the authors try to synthesise the current state of affairs into a taxonomy of models:
\begin{itemize}
	\item \textit{Opaque systems}: these are systems that offer no insight into how their internal workings transform the inputs symbols, usually real-world data, into some output, for example labels that classify the data or predictions of unseen cases. 
		All closed-source algorithms fall under this definition.
	\item \textit{Interpretable systems}: this is the vastest category, as the characteristic of these systems is \textit{transparency} i.e., their inner workings are accessible but the onus of comprehensibility falls completely onto the user.  
		The classical example is that of neural networks, where the mapping from inputs to outputs is inspectable by the user who can, theoretically and depending on her skill, interpret them.
		In the case of neural networks this may be a daunting task due to their distributed and non-symbolic nature; in other classes of machine learning models, for example Bayesian networks, the task may be easier as discussed in Section \ref{sec:explainability-in-bayesian-networks}.
	\item \textit{Comprehensible systems}: systems in this category emit additional symbols together with their outputs with the explicit intent of giving the user the means to interpret and understand the automated decisions.
		The additional symbols may be visualisations, natural language text or any other means of demystifying the output.  
		These extra symbols would need to be graded based on the user's expertise, as comprehension is a property that involves both man and machine.
		An example of such a system would be an image classifier that, together with its output, highlighted the parts of the image it used to make its decision.
	\item \textit{Explainable systems}: the highest level in the taxonomy includes those models that emit an explicit \textit{explanation} i.e., a human-understandable line of reasoning.
\end{itemize}
It is recognised that comprehensibility depends not only on the system's characteristics, but also on the user's ability and knowledge, thus implicitly accepting the view that xAI should include elements from the social sciences.
\textit{Comprehensibility} and \textit{interpretability} are considered separate concepts as comprehension requires transparency but interpretation does not, given that the user may reason only over the emitted extra symbols.
This notion of comprehensibility is expanded into that of \textit{real} explainability, based on a notion of \enquote{ability to formulate, for the user, a line of reasoning that explains the decision making process of a model using human-understandable features of the input data} \citep{Doran2018}.

\section{Importance of Explainability} \label{sec:importance-of-explainability}
As noted by \citet{edwards2018enslaving}, \enquote{businesses and governments are increasingly deploying machine learning (ML) systems to make and support decisions that have a crucial impact on everyday life} so, as \citet{gilpin2018explaining} say, \enquote{it becomes necessary for these mechanisms to explain themselves}.
This feeling of urgency and purpose is echoed throughout the reviewed literature; it seems that even if researchers and the field as a whole cannot agree on a definition of explainability (as discussed in Section \ref{sec:explainability}), there is a keen awareness of the need for models to be explainable.
This intense urge to define explainability and, at the same time, to try and create models exhibiting this property, may be counterproductive as the field risks fragmenting into a series of diverging strands, as noted by \citet{abdul2018trends} and visualised in Figure \ref{fig:xai-citation-network}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.8\textwidth]{literature-review/images/xai-citation-network}}
\caption{Citation network that is emblematic in showing the breadth of research strands in the field of xAI \citep{abdul2018trends}.}
\label{fig:xai-citation-network}
\end{figure}

There are a myriad of reasons brought forth as a justification for the development of explainable AI, and we will review these in the following paragraphs, but it seems timely to start with one in particular: the need introduced by the European Union's broad General Data Protection Regulation (GDPR).
The GDPR was approved by the European Parliament in 2016 and came into effect in 2018.
More than one author cites an urgency to conform to this regulation (for example \citet{doshi2017towards}, \citet{gilpin2018explaining}), most likely referring to Article 22 of the regulation that, supposedly, mandates for a \enquote{right to an explanation} from algorithms.
While algorithmic explainability is undoubtedly a commendable objective, it may be the case that this particular reason to strive for it be a false one.
\citet{edwards2018enslaving} posit that Article 22 of the GDPR actually does not contain the publicised right to an explanation but is \enquote{merely a right to stop processing unless a human is introduced to review the decision on challenge} and as the authors point out, there are, nowadays at least, very few systems without a human in the loop.
Secondly, there is no mandate for the \enquote{explanation} to be understandable by humans, so the result obtained might actually be no explanation at all.
If the analysis of \citet{edwards2018enslaving} were correct, then the urgency advocated by many researchers on the grounds of having to conform to the GDPR would, in reality, turn out to be based on no valid reason.

A second motive brought forward for the need for explainability, for example by \citet{gilpin2018explaining} and \citet{abdul2018trends}, is that comprehensible models are much more likely, or even necessary, to engender users' trust.
While this may very probably be the case, no motives are given for why this should be the purpose of explainable AI and not just a desirable by-product of obtaining explainability. 
A reasoning for why trust may stem from explainability can be found in \citep{Kyrimi2016} where it is claimed that \enquote{the lack of trust may be due to the difficulty of understanding how a prediction is inferred from the given data. As Aristotle wrote `we do not have knowledge of a thing until we have grasped its why, that is to say, its explanation'. Hence, explaining a model's reasoning - its inference - could increase trustworthiness.} 
What could be assumed, is that \textit{trust is a prerequisite for explainability} because if the human does not believe the outputs of the machine, there is little hope for an explanation to come into being.

Other authors, for example, \citet{doshi2017towards} and \citet{guidotti2018survey}, frame the issue as one of moral necessity.
One need not look far to find examples of ML models displaying \textit{covert bias} or making decisions we would regard as unethical; a more in-depth investigation would reveal that this was the case even before the popularisation of \enquote{black box} models as are deep neural networks.
\citet{guidotti2018survey} give a reasonably comprehensive list of classic cases that show the risks inherent in not having comprehensible AI.
The oldest of these dates back to the 1970s and 1980s and tells of a system used to screen job applicants that was still seen to discriminate against minorities, even though programmed to ignore people's ethnicity.
In the same vein but much more recently, the American Military discovered that their computer vision system, developed to differentiate between enemy and friendly tanks automatically, had poor accuracy because it had learned to use the background information of the test set photos instead of the pixels representing the actual vehicles.
Both these cases exemplify how an algorithm may make \enquote{wrong} inferences based on spurious or latent information that was already in the data set but that no human could have imagined being relevant.
Other failures epitomise how a model may learn our own social prejudices; for example, a recent Princeton study \citep{caliskan2017semantics} proved how models trained on web text corpora showed marked bias (towards race, gender ...) that reflected the ones present in our own society.
Based on their findings, the authors went so far as to suggest that transparency would not be enough to uproot bias, since the very \textit{semantics} of language reflects prejudice latent in our culture.

The driving motivation and sense of urgency present in all of the reviewed literature is quite certainly tied to the renewed interest and applicability of AI.
\citet{Preece2018} claims that the interest for explainability is naturally linked to that in AI; if this were true, then it would confirm that a need for transparency is implicit in the field itself.
This would validate the assertion made by \citet{doshi2017towards}, that the need for an explanation stems from an \textit{incompleteness in the formalisation}.
What is meant by this is that optimising for certain objectives may introduce an \textit{unquantifiable bias} into the system, which is very different from mere \textit{uncertainty}.
Mere uncertainty can be rigorously quantified, formalised and reasoned upon by using probability theory; unquantifiable bias is the result of an \textit{incompleteness in formalisation} of the problem that the ML system is tasked with.
For example, this is the case when a system is coded to pursue soft objectives such as \textit{ethics} as such an end-goal may be too abstract and nebulous to be completely formalised.
It might also be the case that a particular objective is far too complex for all its possible outcomes to be exhaustively enumerated.
A good characterisation of objectives leading to incompleteness could be given by applying the concept of \textit{wicked problem} introduced by \citet{Rittel1973} when analysing the issues arising in social planning.
The first defining characteristic of a wicked problem is that the mere definition of the problem in hand is the wicked problem itself, since even its description is dependent on one's idea for solving it; there is no definite locus that one can point to as the source of the problem. 
Other defining features are the absence of \textit{stopping and objective evaluation criteria} and the fact that each solution to a wicked problem is essentially \enquote{one-shot} and unique.  
Even the set of possible solutions and causes are not predetermined and non-stationary. 
The authors noticed how these problems presented a series of defining characteristics, that they were able to generalise into the notion of \textit{wicked problem}.
A classic example are the issues that arise when trying to solve a social planning problem: various parties will have competing objectives and different ideas to obtain them, so no unique best solution is possible; problems are \enquote{at best, only re-solved} \citep{Rittel1973}. 
These situations generally tend to arise when dealing with human values, as there is always a degree of ethical relativism that makes it difficult to develop and evaluate solutions.  
\citet{Lipton2016} is also aware of this and states that \enquote{the demand for interpretability arises when there is a mismatch between the formal objectives of supervised learning and the real world costs in a deployment setting}.
In the presence of such unquantifiable uncertainty, unavoidable in many of the important applications of AI, requiring the resulting model to be open to inspection would make the \enquote{gaps in problem formalization visible to us} and thus enable us to apply our best human judgement to evaluate them and their consequences \citep{doshi2017towards}.
No wicked problem has a solution that is either true or false, but only good or bad; necessarily, the only judge for this can be a human.

\section{Evaluation of Explainability} \label{sec:evaluation-of-explainability}
As concluded in Section \ref{sec:importance-of-explainability}, the fact that ML models operate on incomplete assumptions makes it a necessity to have some form of evaluation of their performance.
As \citet{Lipton2016} states, \enquote{it turns out that many situations arise when our real world objectives are difficult to encode as simple real-valued functions} and this could lead to evident difficulties in optimisation with respect to soft concepts such as ethics and legality which are, however, of paramount importance.
Being able to evaluate an automated explanation lets us \enquote{serve those objectives that we deem necessary but struggle to model formally}.

Unfortunately, the finding outlined in Section \ref{sec:explainability} of there being no consensus on the definition of explainability also necessarily entails that there is no agreed-upon methodology to evaluate such a property.
\citet{doshi2017towards} note as much when they comment \enquote{unfortunately, there is little consensus on what interpretability in machine learning is and how to evaluate it for benchmarking}.
This makes perfect sense because trying to evaluate something without first having defined it, is worse than trying to hit a moving target.
Once again, the feeling shared among many authors is that \enquote{inmates are running the asylum}.

Some authors have tried to put some order in the barrage of proposed methods with \citet{doshi2017towards} providing one of the most compelling attempts.
These authors set out to outline a taxonomy, having noted a \enquote{lack of rigour} and how current interpretability approaches usually fall into two categories: \textit{interpretability in the context of an application and interpretability via a quantifiable proxy}.
 The former approach assumes that \enquote{if the system is useful in either a practical application or a simplified version of it, then it must be somehow interpretable}; the latter sees researchers claim that a model class is interpretable and then present algorithms to optimise within that class.
 In their words, both classes rely on a notion of \enquote{you'll know it when you see it}.
 The taxonomy the authors propose is laid out in Figure \ref{fig:xai-taxonomy} and borrows from methods already standard in human-computer interaction and visualisation; the guiding ideal is that \enquote{evaluation of applied work should demonstrate success in the application} and thus the best kind of evaluation is the one that involves humans the most:
 \begin{itemize}
  \item \textit{Functionally-grounded Evaluations}: at the lowest level of their taxonomy are those methods requiring no human-in-the-loop and that evaluate the quality of an explanation given by a system by using some proxy measure; the advantage is the low cost, but the tradeoff is a lack of specificity.
 A proxy measure that has already been human-validated as regards its explainability, for example a decision tree, a set of rules or a linear model \citep{guidotti2018survey}, may be substitute to estimate the explainability of more complex systems.
 \item \textit{Human-grounded Evaluation}: the second level of evaluation involves humans, albeit not domain expert ones, carrying out simplified versions of the target application; this kind of setup enables the testing of more general notions of explainability.
 \item \textit{Application-grounded evaluation}: this is considered the gold standard evaluation; the authors claim that there is no better way to evaluate explainability that having a domain expert test it in the context of a real task.
In their words, \enquote{the best way to show that the model works is to evaluate it with respect to the task}: \enquote{for example, a visualization for correcting segmentations from microscopy data would be evaluated via user studies on segmentation on the target image task; a homework-hint system is evaluated on whether the student achieves better post-test performance.  Specifically, we evaluate the quality of an explanation in the context of its end-task, such as whether it results in better identification of errors, new facts, or less discrimination}.
\end{itemize}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{literature-review/images/xai-taxonomy}}
\caption{Taxonomy of methods for the evaluation of explanations \citep{doshi2017towards}.}
\label{fig:xai-taxonomy}
\end{figure}

\citet{guidotti2018survey} outline a taxonomy developed along a different axis; specifically they identify: 
\begin{itemize}
  \item \textit{methods to explain black box models};
  \item \textit{methods to explain black box outcomes};
  \item \textit{methods to inspect black boxes and methods to design transparent boxes}.
\end{itemize}
An essential factor that the authors identify is that \textit{evaluation is a graded notion}; as also noted by \citet{gilpin2018explaining}: different users, with different expertise and background, may rate the quality of the same explanations very differently.
Another point they make - which is quite novel - is that \textit{time should also be part of an evaluation}: depending on time available a user may prefer a more straightforward or more elaborate explanation.
They end by noting that \textit{very few studies take user expertise and the time taken to understand the proposed explanation into account}.
The concept of \textit{time to understand an explanation} will be part of the methods of this thesis.
